Effectful Semantics for Substructural SSA

### 2023-11-03 pldi24/paper.tex
more intro

@@

\begin{abstract}
  TODO THIS
\end{abstract}

@@

\TODO{abstract?}

@@

Static Single Assignment, or SSA, is a cornerstone of modern compiler design, on which projects such as the LLVM compiler infrastructure, GCC's GIMPLE intermediate representation, Cranelift, and MLIR are based. One of the core advantages of putting programs in SSA form is the ability to reason about programs in a more equational manner: since variables always have exactly one definition, we can justify complex optimizations like global value numbering (which would otherwise need to take into account complex flow-sensitive definitions) and more easily implement various analyses.

\TODO{text...}

>>>>

\newcommand{\todo}[1]{\stepcounter{todos} \textcolor{red}{TODO \arabic{todos}:} #1}

@@

\begin{abstract}
  Static single-assignment form (SSA) is one of the most widely used
  compiler intermediate representations. In this paper, we build on
  the observation that SSA can be seen as a language of basic blocks
  with arguments (or first-order procedures tail-calling one another)
  to give a type theory for SSA programs. Notably, the type system and
  equational theory of our language is designed to support
  substituting impure terms for variables, which makes it easy to
  validate compiler optimisations such as hoisting assignments out of
  loops.

  We give a categorical axiomatisation of our type theory and show
  that our equational theory is sound with respect to it. Our
  categorical semantics generalises the notions of premonoidal
  category and Freyd category to support substructural features,
  including both support for substructural types, as well as
  F\"{u}hrmann's notion of central, copyable and discardable
  morphisms.

  Finally, we exhibit a number of concrete models of our categorical
  axiomatisation, including in particular a model of TSO-style weak
  memory based on the semantics of \citet{sparky}. This demonstrates
  that our approach is strong enough to connect machine models of
  concurrency to compiler IRs supporting a rich collection of sound
  rewrites.
\end{abstract}

@@

@@

Static Single Assignment, or SSA, has been a cornerstone of compiler
design ever since its introduction by \cite{ssa-original} in the late
1980s. Older compiler intermediate representations permitted the same
local variable to be assigned multiple times, whereas SSA requires
each variable bindings to be immutable -- they cannot be modified
after initialisation. This makes many dataflow analyses easier to
compute, and (just as in functional languages) makes it much easier to
justify compiler optimisations in terms of equational rewrites.

In fact, the relationship between SSA form and functional programming
has been known for decades: \citet{kelsey-ssa-cps} showed that SSA can
be seen as a subset of continuation-passing style, and
\citet{appel-ssa} built on this to observe that every SSA program can
be viewed as a collection of first-order functions which tail-call one
another. This makes the equivalence of the functional view to the
traditional $\phi$-node presentation very easy to see. In the
functional style, each tail call lists its arguments. In contrast, a
$\phi$-node collects together the arguments from every callsite, which
lets the tail-calls/jumps to omit their arguments. Thus, the main
difference between SSA and a functional program is where in the
program text the arguments to a tail call are listed.

\TODO{Example here?}

This ``basic blocks with arguments'' presentation of SSA has become
very popular in industry, with compiler backends such as
Cranelift~\cite{cranelift}, MLIR~\cite{mlir}, and SIL~\cite{SIL} all
using this representation. This tight connection might suggest that
the semantics of SSA representations is trivial: we can just interpret
an SSA program as a corresponding functional program, and then rewrite
the SSA programs using the equational theory of functional programs.

While this is a perfectly sound thing to do, it is not strong enough
to write a good compiler. Compiler IRs like SSA are fundamentally
about effectful programs, and the equational theory of effectful
functional programs is very weak. Instead of a full substitution rule,
the equational theory of impure functional languages only justifies
substituting values for variables (the $\beta$-value rule). This weak
substitution rule suffices to validate some basic compiler
optimisations such as constant propagation and copy propagation, but
is not strong enough to justify many other essential compiler
optimisations. For example, hoisting a store out of a loop or
conditional involves moving an effectful term (an assignment) out of
its original position in the program. As another example, dead store
elimination removes an effectful term from the program altogether. \todo{Another example?}

These examples illustrate that many optimisations require taking
\emph{effectful} terms and moving them around and substituting them
for variables.

% 
% , on which projects such as the LLVM compiler infrastructure, GCC's GIMPLE intermediate representation, Cranelift, and MLIR are based. One of the core advantages of putting programs in SSA form is the ability to reason about programs in a more equational manner: since variables always have exactly one definition, we can justify complex optimizations like global value numbering (which would otherwise need to take into account complex flow-sensitive definitions) and more easily implement various analyses. 
% 
\TODO{text...}
